---
title: "EDA on Instacart Database"
author: "Gaurang Swarge"
date: "2/15/2019"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)

options(width = 100)


#EDA on the data

library(plyr)
library(data.table)
library(dplyr)
library(ggplot2)
library(stringr)
library(DT)
library(scales)
library(knitr)

#Reading all the Datasets
orders <- fread("Dataset/orders.csv")
products <- fread("Dataset/products.csv")
departments <- fread("Dataset/departments.csv")
aisles <- fread("Dataset/aisles.csv")
opp <- fread("Dataset/order_products__prior.csv")
opt <- fread("Dataset/order_products__train.csv")


#converting to factor variables
orders$order_hour_of_day <- as.numeric(orders$order_hour_of_day)
orders$order_dow <- as.factor(orders$order_dow)
orders$eval_set <- as.factor(orders$eval_set)

products$product_name <- as.factor(products$product_name)

departments$department <- as.factor(departments$department)

aisles$aisle <-  as.factor(aisles$aisle)


#Merging the tables:

products_aisles <- merge(products,aisles,by="aisle_id")
products_aisles_dept <- merge(products_aisles,departments,by="department_id")


```

### Top 50 most Ordered Products
```{r top50ordered}

temp <- orders  %>%
  inner_join(opp,by="order_id") %>%
  inner_join(products,by="product_id")
 
  summary(temp)  
  kable(head(temp,15))
  highest_ordered_product <- sort(table(temp$product_name),decreasing = TRUE)
  highest_ordered_product <- highest_ordered_product[1:50]
  kable(head(highest_ordered_product),15)
  #plot(highest_ordered_product)
  hop <-as.data.frame(highest_ordered_product)

hop %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none" )+
    scale_y_continuous(breaks=seq(50000, 480000, 25000))+
    labs(title="Highest Ordered Products - Top 50", y= "No of Orders")
  
```

### Top 50 most Re-Ordered Products
```{r }
temp <- orders  %>%
  inner_join(opp,by="order_id") %>%
  inner_join(products,by="product_id") %>%
  filter(reordered==1)

summary(temp)  
kable(head(temp,15))
highest_reordered_product <- sort(table(temp$product_name),decreasing = TRUE)
highest_reordered_product <- highest_reordered_product[1:50]
kable(head(highest_reordered_product),15)
#plot(highest_ordered_product)
hrop <-as.data.frame(highest_reordered_product)

hrop %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
  geom_bar(stat="identity")+
  theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none")+
  #coord_cartesian(ylim=c(1000,20000,1000))+
  scale_y_continuous(breaks=seq(50000, 480000, 25000))+
  labs(title="Highest re-Ordered Product - Top 50", y="No of Orders")
```


### Finiding unique products which are not reordered
Organic Italian Parsley Bunch & Blueberries are 2 products which are part of highest ordered products but are not part of the top 50 re-ordered products

``` {r prodnotrepordered}
uniq_prod <- hop %>%
  anti_join(hrop, by="Var1")
  #data.frame(unique_prod=union(hop$Var1, hrop$Var1))
uniq_prod
```
 
### Least ordered products
 
 
``` {r leastordered}


bottom25_products <- orders  %>%
  inner_join(opp,by="order_id") %>%
  count(product_id) %>%
  arrange(desc(n)) %>%
  tail(25) %>%
  merge(products,by='product_id')
kable(head(bottom25_products,25))
  
bottom25_products %>% ggplot(aes(x=product_name,y=n, fill=product_name)) +
  geom_bar(stat="identity")+
  theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none")+
  #coord_cartesian(ylim=c(1000,20000,1000))+
  scale_y_continuous(breaks=seq(0, 5, 1))+
  labs(title="Least Ordered Products - Last 25", y="No of Orders")
```
 
### Order hour of the day vs number of orders of customers

To find out what time of the day most orders are placed

```{r orderhourofday}
  p<-orders %>% 
    group_by(order_hour_of_day,order_number)%>%
    summarise(count = n()) %>%
    arrange(desc(count))
  
  kable(head(p,10))
  #dim(p)
  
  p %>% ggplot(aes(x=order_hour_of_day,y=order_number,fill=order_number))+
    geom_tile(aes(fill=count),colour = "white") + 
    scale_fill_gradient(low = "aquamarine",high = "red")+
    theme(axis.text.x=element_text(angle=60, hjust=1),panel.grid.major = element_line(colour = "lightgrey"))+
    scale_x_continuous(breaks=seq(0, 24, 1))+
    scale_y_continuous(breaks=seq(0, 100, 5))+
    labs(title="Order Hour of Day", x="Order Hour of Day", y="Count of order numbers")

```
### 2. Order Day of the week vs the Order Hour of the day

```{r orderdayofweek}
temp4 <- orders %>%
    mutate(order_dow = as.factor(order_dow)) %>%
    mutate(order_hour_of_day = as.factor(order_hour_of_day)) %>%
    group_by(order_dow,order_hour_of_day) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
    
  kable(head(temp4),10)
  
  temp4 %>% ggplot(aes(x=order_dow,y=order_hour_of_day))+
    geom_tile(aes(fill=count),colour = "white") + 
    scale_fill_gradient(low = "aquamarine",high = "red")+
    #scale_fill_distiller(palette = "Spectral")+
    theme(axis.text.x=element_text(angle=60, hjust=1),panel.grid.major = element_line(colour = "grey"))+
    labs(title="Which Day and what time do people order the most?",x="Day of the Week", y="Order Hour of the day")
  
```


### 3. After How many days do customers place a Re-Order?
7th Day and the 30th day is when people reorder the most<br/>

```{r reorderdays}
 p<-orders 

 p %>% ggplot(aes(x=days_since_prior_order,fill=order_dow))+
    geom_histogram(stat = "count")+
    theme(axis.text.x=element_text(angle=60, hjust=1),panel.grid.major = element_line(colour = "grey"))+
    scale_x_continuous(breaks=seq(0, 30, 1))+
    labs(x="No. of Days since last Order",y="Count of Orders")
```
 
### 4. Finding out total %ge of products which are reordered
As you can se about 60% products are generally reordered & 40% of Products are not ordered again<br/> 

```{r percentreordered}
temp2 <- opp %>%
    group_by(reordered) %>%
    summarise(count = n()) %>%
    mutate(reordered = as.factor(reordered)) %>%
    mutate(percentage = count/sum(count))
  
  kable(head(temp2,10))
  
  temp2 %>% ggplot(aes(x=desc(reordered),y=percentage, fill=reordered))+
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),panel.grid.major = element_line(colour = "grey"))+
    scale_x_continuous(breaks=seq(0, 1, 1))+
    scale_y_continuous(breaks=seq(0, 1.0, .15))+
    labs(title="%ge of products Reordered", x="Reordered or Not")
```

### Maximum Product varieties in each Department
'Personal Care' and 'Snacks' has the maximum product varieties<br/>

```{r prodvaritiesdept}
temp3 <- products %>%
    group_by(department_id) %>%
    summarise(count = n()) %>%
    arrange(desc(count)) %>% 
    left_join(departments,by="department_id")%>%
    top_n(20)
  
  kable(head(temp3),10)
  
  temp3 %>% ggplot(aes(x=reorder(department,-count),y=count, fill=department))+
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"))+
    labs(title="Max number of product variety in which Department? (Top 20)",y="Count of Products")
  

```

### 5. Products from which Departments are ordered the most ?
```{r highestordereddepts}
  temp6 <- opp %>% 
   # arrange(desc(-order_id)) %>% 
    left_join(products,by="product_id")%>%
    left_join(departments,by="department_id")%>%
    select(order_id,department)
    #distinct(order_id,department)
  
  kable(head(temp6,10))
  dim(temp6)

  hod <-  sort(table(temp6$department),decreasing = TRUE)
  dim(hod)
  hod <- hod[1:21]
  kable(hod)
  hod <-as.data.frame(hod)
  
  hod %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none" )+
    scale_y_continuous(breaks=seq(30000, 9500000, 500000))+
    labs(title="Highest Ordered Department", y= "No of Orders")

```

### 6. Products from which from which  Departments are reordered the most?
```{r highestreordereddepts}
  temp6 <- opp %>% 
    left_join(products,by="product_id")%>%
    left_join(departments,by="department_id")%>%
    filter(reordered == 1) %>%
    select(order_id,department)
  
  kable(head(temp6,10))
  dim(temp6)
  
  hrod <-  sort(table(temp6$department),decreasing = TRUE)
  dim(hrod)
  hrod <- hrod[1:21]
  kable(hrod)
  hrod <-as.data.frame(hrod)
  
  hrod %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none" )+
    scale_y_continuous(breaks=seq(30000, 9500000, 500000))+
    labs(title="Highest Re-Ordered Department", y= "No of Orders")
```

#### 7. Change in the Department rank during the reorder
Beverages  jump to 3rd position compared to 4th during the first order <br/>
Bakery jumps to 6th position during reordered compared to 7th in the first order <br/>
Deli jumps to 8th position during  reorder compared to 9th in the first order<br/>
Similarly 'Breakfast' & 'meat seafood' also jumps up one position during the reorder compared to first time order
```{r changedeptrank}
  uniq_dept <- hod %>%
    left_join(hrod, by="Var1")%>%
    arrange(desc(Freq.x))
  uniq_dept
```

### Maximum Product varieties in each Aisles
'Personal Care' and 'Snacks' has the maximum product varieties<br/>

```{r prodvaritiesaisle}
temp3 <- products %>%
    group_by(aisle_id) %>%
    summarise(count = n()) %>%
    arrange(desc(count)) %>% 
    left_join(aisles,by="aisle_id")%>%
    mutate(aisle = if_else(as.character(aisle) %in% c('missing','other'), 'G-Others',as.character(aisle))) %>% 
    top_n(50)
  
  kable(head(temp3),10)
  
  temp3 %>% ggplot(aes(x=reorder(aisle,-count),y=count, fill=aisle))+
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none")+
    labs(title="Max number of product variety in which Aisles? (Top 50)",y="Count of Products")
  

```

### Top 50  highest ordered Aisles?
```{r highestorderedaisles}

temp6 <- opp %>% 
    left_join(products,by="product_id")%>%
    left_join(aisles,by="aisle_id")%>%
    select(order_id,aisle)
  
  kable(head(temp6,10))
  dim(temp6)
  
  hoa <-  sort(table(temp6$aisle),decreasing = TRUE)
  dim(hoa)
  hoa <- hoa[1:50]
  kable(head(hoa,10))
  hoa <-as.data.frame(hoa)
  
  hoa %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none" )+
    scale_y_continuous(breaks=seq(300000, 3700000, 500000))+
    labs(title="Highest Ordered Aisles - Top 50", y= "No of Orders")
```

### Top 50 highest reordered Aisles?
```{r highestreorderedaisles}
temp6 <- opp %>% 
    left_join(products,by="product_id")%>%
    left_join(aisles,by="aisle_id")%>%
    filter(reordered == 1)
    #select(order_id,aisle)
  
  kable(head(temp6,10))
  dim(temp6)
  
  hroa <-  sort(table(temp6$aisle),decreasing = TRUE)
  dim(hroa)
  hroa <- hroa[1:50]
  kable(hroa)
  hroa <-as.data.frame(hroa)
  
  hroa %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none" )+
    scale_y_continuous(breaks=seq(100000, 2700000, 500000))+
    labs(title="Highest RE-Ordered Aisles - Top 50", y= "No of Orders")
```

### Unique Aisles which are not reordered again
'oils vinegars', condiments, 'spices seasonings' are three aisles which are not reordered within 30 days

```{r uniqueaisles}
uniq_aisles <- hoa %>%
    anti_join(hroa, by="Var1")
  uniq_aisles
  
```

### Looking into Customers - Highest Ordered products in the First Order 
```{r highestorderedfirstorder}
temp <- orders  %>%
  inner_join(opp,by="order_id") %>%
  inner_join(products,by="product_id")%>%
  filter(order_number<2)%>%
  select(user_id,order_number,product_id,product_name)
  #filter(add_to_cart_order == 1)
  
  summary(temp)  
  kable(head(temp,15))
  hopfo <- sort(table(temp$product_name),decreasing = TRUE)
  hopfo <- hopfo[1:50]
  kable(head(hopfo,15))
  dim(hopfo)
  #plot(highest_ordered_product)
  hopfo <-as.data.frame(hopfo)
  
  hopfo %>% ggplot(aes(x=Var1,y=Freq, fill=Var1)) +
    geom_bar(stat="identity")+
    theme(axis.text.x=element_text(angle=60, hjust=1),axis.title.x = element_blank(),panel.grid.major = element_line(colour = "grey"),legend.position="none")+
    #coord_cartesian(ylim=c(1000,20000,1000))+
    scale_y_continuous(breaks=seq(5000, 30000, 5000))+
    labs(title="Highest Ordered Products in First Order - Top 50", y="No of Orders")
```

### Which products out of the highest ordered products, are not part of the first order of the customer
These 4 products inspite of being part of highest ordered products, are not ordered during the first order by the customers
1     Organic Whole String Cheese 59676
2      Organic Granny Smith Apple 58779
3 Organic Unsweetened Almond Milk 57895
4                    Spring Water 56087
```{r }
First_prod <- hop %>%
    anti_join(hopfo, by="Var1")
  #data.frame(unique_prod=union(hop$Var1, hrop$Var1))
First_prod
```

### Avg no of products in the first order of the customer
Mean   :10.08  means, on an average customer purchases 10 products in the first order<br/>
```{r avgprodfirstorder}

temp <- orders  %>%
  inner_join(opp,by="order_id") %>%
  inner_join(products,by="product_id")%>%
  filter(order_number<2)%>%
  group_by(user_id,order_number)%>%
  summarise(cnt = n())
  #select(user_id,order_number,product_id,product_name)

summary(temp)  
kable(head(temp,15))
```


```{r }


```


```{r }

```


```{r }

```


```{r }

```


```{r }

```


```{r }

```


```{r }

```
